---
title: "ps5-gapminderData"
author: "Lydia Sorbo"
date: "2023-02-17"
output: html_document
---

## 1. **Load and Check Data**

**Question 1.2**

```{r message=FALSE}
library(tidyverse)
```

```{r}
gapminder <- read_delim("gapminder.csv.bz2")
```

The gapminder data has `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.

**Question 1.3**

```{r}
gapminder %>% 
  sample_n(5)
```

## 2. **Descriptive Statistics**

**Question 2.1**

There are `r length(unique(gapminder$name))` countries in the dataset.

There are `r length(unique(gapminder$iso2))` iso2 codes in the dataset.

There are `r length(unique(gapminder$iso3))` iso3 codes in the dataset.

**Question 2.2.a**

```{r}
gapminder %>% 
  group_by(iso2) %>% 
  reframe(totalCountries = length(unique(name)), unique(name)) %>% 
  tail(5)
```

The iso2 code for Namibia is being recognized as missing since it is "NA", resulting in 249 iso2 codes for 250 countries.

**Question 2.2.b**

```{r}
gapminder %>% 
  group_by(iso3) %>% 
  reframe(totalCountries = length(unique(name)), unique(name))
```

The iso3 codes CHANISL, NLD CURACAO, and GBM have missing country names, resulting in 253 iso3 codes for 250 countries.

**Question 2.3**

```{r}
gapminder %>% 
  filter(!is.na(time)) %>% 
  summarize(minYear = min(time), maxYear = max(time))
```

The minimum year is 1960 and the maximum year is 2019.

## 3. **CO2 Emissions**
