---
title: "ps5-gapminderData"
author: "Lydia Sorbo"
date: "2023-02-17"
output: html_document
---

## 1. **Load and Check Data**

**Question 1.2**

```{r message=FALSE}
library(tidyverse)
```

```{r}
gapminder <- read_delim("gapminder.csv.bz2")
```

The gapminder data has `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.

**Question 1.3**

```{r}
gapminder %>% 
  sample_n(5)
```

## 2. **Descriptive Statistics**

**Question 2.1**

There are `r length(unique(gapminder$name))` countries in the dataset.

There are `r length(unique(gapminder$iso2))` iso2 codes in the dataset.

There are `r length(unique(gapminder$iso3))` iso3 codes in the dataset.

**Question 2.2.a**

```{r}
gapminder %>% 
  group_by(iso2) %>% 
  reframe(totalCountries = length(unique(name)), unique(name)) %>% 
  tail(5)
```

The iso2 code for Namibia is being recognized as missing since it is "NA", resulting in 249 iso2 codes for 250 countries.

**Question 2.2.b**

```{r}
gapminder %>% 
  group_by(iso3) %>% 
  reframe(totalCountries = length(unique(name)), unique(name))
```

The iso3 codes CHANISL, NLD CURACAO, and GBM have missing country names, resulting in 253 iso3 codes for 250 countries.

**Question 2.3**

```{r}
gapminder %>% 
  filter(!is.na(time)) %>% 
  summarize(minYear = min(time), maxYear = max(time))
```

The minimum year is 1960 and the maximum year is 2019.

## 3. **CO2 Emissions**

**Question 3.1**

```{r}
gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2)) %>% 
  filter(is.na(co2_PC)) %>% 
  reframe(missingco2 = length(co2), missingco2PC = length(co2_PC)) %>% 
  arrange(desc(missingco2)) %>% 
  head(5)
```

2017-19 have the most missing emissions data at 217.

**Question 3.2**

```{r}
gapminder %>% 
  filter(!is.na(co2)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | 
           iso3 == "IND", iso3 == "CAN", iso3 == "RUS") %>% 
  ggplot(aes(time, co2, col=factor(iso3))) +
  geom_line() +
  labs(x = "Time (years)", y = "CO2 Emissions", 
       col = "Country")
```

**Question 3.3**

```{r}
gapminder %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | 
           iso3 == "IND", iso3 == "CAN", iso3 == "RUS") %>% 
  ggplot(aes(time, co2_PC, col=factor(iso3))) +
  geom_line() +
  labs(x = "Time (years)", y = "CO2 Per Capita Emissions", 
       col = "Country")
```
